Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic

### 2021-08-31 write-up/covid-writeup.tex
Updated IFR to match Raftery

\begin{abstract}
  Coronavirus case-count data has influenced government policies and drives most epidemiological forecasts. Limited testing is cited as the key driver behind minimal information on the COVID-19 pandemic. While expanded testing is laudable, measurement error and selection bias are the two greatest problems limiting our understanding of the COVID-19 pandemic; neither can be fully addressed by increased testing capacity. In this paper, we demonstrate their impact on estimation of point prevalence and the effective reproduction number. We show that estimates based on the millions of molecular tests in the US has the same mean square error as a small simple random sample.  To address this, a procedure is presented that combines case-count data and random samples over time to estimate selection propensities based on key covariate information. We then combine these selection propensities with epidemiological forecast models to construct a \emph{doubly robust} estimation method that accounts for both measurement-error and selection bias.  This method is then applied to estimate Indiana's prevalence using case-count, hospitalization, and death data with cumulative demographic information, one of United States's only two statewide random molecular sample collected from April 25--29th, and Facebook's COVID-19 symptom survey.  We end with a series of recommendations based on the proposed methodology.
\end{abstract}

@@

Due to monetary and time constraints, however, random testing is performed infrequently.  As of June 2021, to the best of our knowledge, Indiana and Ohio are the only states to conduct statewide random sample testing\footnote{This is the only random sample to collect both seroprevalence and diagnostic testing results. The CDC has conducted several seroprevalence studies. See Section~\ref{section:testinginfo} for further discussion.}, which was collected from April 25--29, 2020~\citep{Yiannoutsos2021}.  Such infrequent random testing is likely to provide insufficient information to help researchers and policy makers better understand the disease trajectory which can change rapidly over time.  Therefore, while random testing may be preferable in theory, in practice governments, researchers, and policy makers continue to use coronavirus case-counts to understand the impact of COVID-19 on the population and make data-informed decisions.

@@

 This article discusses the relationship between three statistical concepts: selection bias, measurement error, and population size.  While the impact of measurement error~\citep{Smeden2019} and selection bias~\citep{Keiding2016} on estimation are both well-studied topics, here we provide a new perspective by building on the work of \cite{Meng2018} who studied an error decomposition to understand the relationship between selection bias and population size.  Speficially, we quantify the interaction between measurement-error and selection bias on statistical error, showing how the sign and magnitude can change drastically.
 % the impact on the effective sample size can therefore be quite large.

@@

 One core component of coronavirus research is epidemiological state-space modelling of case-count and death data. These models can be used to answer a variety of research questions including case-count forecasting, estimation of the effective reproduction number, and estimation of quarantine and other health policies on infectious disease dynamics.  The basic approach is a deterministic compartmental model called the susceptible-infectious-recovered (SIR) model.  A probabilistic extension was proposed by~\cite{Osthus2017} to model one-dimensional time series of infected proportions. \cite{Song2020} extended this approach to incorporate interventions and assess interventions on COVID-19 epidemic in China. \cite{Hao2020} builds an extended SEIR model (SAPHIRE) to account for various presymptomatic infectiousness, time-varying ascertainment rates, transmission rates and population movements.

@@

 Between April 25--29, 2020, Indiana conducted statewide random molecular testing of persons ages $\geq 12$ years to assess prevalence of active infection to SARS-CoV-2~\citep{Yiannoutsos2021}. A stratified random sampling design was conducted using Indiana’s 10 public health preparedness districts as sampling strata. 15,495 participants were contacted resulting in a final sample size of 3,658. Demographic data was collected (e.g., summary statistics on age, sex, and race) as well as data on whether they experienced any COVID-19 compatible symptoms during the past 2 weeks or had shared a household with someone who had a positive test result for SARS-CoV-2. During May 2--3, 2020, an additional nonrandom sample of 898 individuals was also collected.
 % In this paper, the nonrandom sample is used to assess whether the cumulative summary statistics are appropriate in constructing the inverse-probability weights as well as in constructing appropriate weights using symptom information.

@@

 Here we assume access to a probability sample measuring both the outcome and exact same set of covariates.  Let $\tilde I_j$ be an indicator that the individual was included in the probability sample and $\tilde W_j$ be the probability of inclusion for the $j$th individual (i.e., equal to $n/N$ for a simple random sample of size $n$). In recent work,~\cite{Chen2019} use a probability sample to construct a design-unbiased estimator of the second term
 \begin{equation}
 \label{eq:auxinfoprob}
 \sum_{j=1}^N I_j \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right)  + \sum_{i=1}^N \tilde I_j \tilde W_j \log ( 1 - \pi (X_j; \theta)).
 \end{equation}
 Taking an expectation with respect to the sampling design yields~\eqref{eq:propensity}. Solving~\eqref{eq:auxinfoprob} can be done by iteratively re-weighted least squares (IRLS); see Section~\ref{app:irls} in the suppementary materials for algorithmic details.

@@

 If sufficiently large random samples are collected at each time $t =1,\ldots,T$, then the pseudo-likelihood can be re-written as in~\eqref{eq:auxinfoprob} and propensities estimated separately per time point.  Unfortunately for COVID-19, large probabilistic samples are not available at every time within a given region.  Here we consider a non-parametric kernel-based approach where the selection propensity at time~$t$, denoted $\hat \theta_t$, maximizes the smoothed pseudo-likelihood
 $$
 \sum_{t^\prime=1}^T K_h(|t^\prime - t|) \left[ \sum_{j=1}^N I_{j,t^\prime} \log \left( \frac{\pi_t (X_{j,t^\prime}; \theta)}{1-\pi_t(X_{j,t^\prime}; \theta)} \right) + \sum_{j=1}^N \tilde W_{j,t^\prime} \tilde I_{j,t^\prime} \log \left( 1 - \pi_t (X_{j,t^\prime}; \theta) \right) \right]
 $$
 where $K_h$ is a kernel function with tunable parameter $h$. Given $\pi (x;\hat \theta_t)$, the prevalence estimator $\bar y_{n,t}^\star$ is given by~\ref{eq:ipwest}.

@@

 \begin{lemma}[Variance of IPW estimator] \normalfont
 \label{lemma:ipw}
 % \textcolor{red}{Must check.}
 The estimates~$\hat \mu_t := \bar y_{n,t}^\star$, $\hat \pi_{j,t}$, $\hat FP$, and $\hat FN$ are solutions to the following set of estimating equations
 $$
 \Phi_n (\eta_t) =
 \left (
 \begin{array}{c}
 \frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - FP - (1-FP-FN) \cdot \mu_t}{\pi_{j,t}} \\
 \frac{1}{N} \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
 \frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
 \frac{1}{n_{FN}} \sum_{j \in S_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
 \end{array}
 \right ) = {\bf 0}
 $$
 where $\eta_t = (\mu_t, \pi_t, FP, FN)$. Under certain regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (n^{-1/2})$  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (n^{-1})$ where $V_t^{(IPW)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$ with $E[\cdot]$ and $\text{Var} (\cdot)$ are under the joint randomization of propensity score and sampling designs.
 \end{lemma}

@@

 \subsection{Doubly robust estimation.}
 Rather than relying solely on IPW or epidemiological forecasts, here we combine forecasting and inverse-probability weighting by extending recent work by~\cite{Chen2019} to account time-varying propensities and measurement-error. In this context, the doubly-robust estimator is given by
 $$
 \bar y_{n}^{(DR)} = \frac{1}{N} \sum_{j=1}^N \hat \mu (x_j) + \frac{1}{\sum_{j=1}^N I_j w (x_j)} \sum_{j=1}^N I_j w(x_j) \left( \frac{Y_j - FP}{1 - FP - FN} - \mu(x_j) \right).
 $$
 where the estimates $\hat \mu(x_j)$ are not corrected for measurement-error as they are built to estimate true case count from death-only data. This estimator is called ``doubly-robust'' because it is consistent if either the model-based forecasts or the time-varying propensities are correctly specified. A statistical error decomposition can be derived
 \begin{align}
 \label{eq:statdecomp3}
 \begin{split}
 \rho_{\tilde I (X), Y-\mu(X)} &\times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y-\mu(X)}  \\
 &\times \underbrace{\left[ 1 - \frac{\rho_{\tilde I(X),Y} \sigma_{Y}}{\rho_{\tilde I(X),Y-\mu(X)} \sigma_{Y-\mu(X)}} \times \tilde \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{\tilde f_0 (1-\bar Y) + \tilde f_1 \bar Y} \right]}_{\tilde D_M}.
 \end{split}
 \end{align}
 See Appendix~\ref{app:DRderivation} for the derivation. Comparing~\eqref{eq:statdecomp3} to~\eqref{eq:statdecomp2} shows that the addition of the model-based forecast impacts the estimation error in two ways.  First, if the model is adequate, then one may expect a reduction in the problem difficulty and (potentially) in the data quality components. Note that the model-based forecast does not impact the data quantity component.

 If the model-based forecast is correct then $E_{\bf R} [ \rho_{\tilde I (X), Y-\mu(X)} ] = 0$  and therefore $\rho_{\tilde I(X), Y-\mu(X)} = O(N^{-1})$; however, if the forecasts are not correctly specified then the data quality index is unlikely to inversely scale with population size. Interestingly, the impact of measurement-error on data quality now depends on a relative comparison of the data quality and problem difficulty of the weighted estimator and the doubly-robust estimator.  For decent models, one would expect this ratio to be greater than one.  Therefore, $\tilde \Delta < 0$ implies an expected increase in impact of measurement-error and $\tilde \Delta > 0$ implies the magnitude could shrink toward zero or change sign and have an even large magnitude in the opposite direction.  Again, if the data quality is not at the level of $N^{-1}$, then confidence intervals constructed from a DR estimator are likely to put too much confidence in the sheer size of the data.

@@

 \begin{lemma}[Variance of doubly-robust estimator] \normalfont
 \label{lemma:dr}
 The estimates~$\mu_t := \bar y_{n,t}^\star$ and $\hat \pi_{j,t}$
 are solutions to the following set of estimating equations
 $$
 \Phi_n (\eta_t) =
 \left (
 \begin{array}{c}
 \frac{1}{N} \left[ \sum_{j=1}^N \frac{I_{j,t}}{\pi_{j,t}} \left( \left( \frac{Y_{j,t} - FP}{1-FP-FN} - \hat \mu_{j,t}  \right)  - \mu_t \right) + \frac{1}{\sum_{j=1}^N \tilde I_j \tilde W_j} \sum_{j=1}^N  \tilde I_j \tilde W_j \hat \mu_t (X_{j,t}) \right] \\
 \frac{1}{N} \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
 \frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
 \frac{1}{n_{FN}} \sum_{j \in S_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
 \end{array}
 \right ) = {\bf 0}
 $$
 where $\eta_t = (\mu_t, \pi_t, FP, FN)$. Under regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (n^{-1/2})$ and $\text{var} (\hat y_{n,t}) = V_{t}^{(DR)} + o (n^{-1})$ where $V_t^{(DR)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$ with $E[\cdot]$ and $\text{Var} (\cdot)$ are under the joint randomization of propensity score and sampling designs.
 \end{lemma}

@@

 \item {\bf Random/Nonrandom statewide sample}: Table~\ref{tab:indiana} summarizes data from a random sample from April 25--29th as well as a nonrandom sample obtained between May 2--3 in racial/ethnic minority communities~\cite{Yiannoutsos2021}. The nonrandom sample is not a random subsample of the overall case-counts; however, this data provides important supplementary covariate information.
 % Namely, the samples collected included covariate information on (1) prior symptoms, (2) COVID-19 positive individual in household, and (3) prior COVID-19 positive test.
 To account for the nonrandom sample being from high risk areas, we adjust estimates to match the statewide positivity rate of 11.7\% between May 1--2. These adjusted rates are presented in parentheses in Table~\ref{tab:indiana} along with the relevant covariate information.

@@

 The IPW analysis is next extended to the time-varying setting.  To do so, we make use of the Facebook symptom survey, hospitalization, and testing data.  Recall the COVID-19 testing and positive case counts are known by age, gender, and racial demographics. Symptom status (e.g., fever, coughing, shortness of breath) and related covariate information (COVID-19 positive contact), however, are currently unavailable\footnote{Whether this is due to data privacy considerations or lack of data collection efforts is unclear.}.  As COVID-19 symptom status is likely to alter the testing propensity and likelihood of active infection, for illustrative purposes Facebook symptom survey and hospitalization data are used in two ways to imputate symptom status per strata. We outline the imputation methods below; see Section~\ref{section:discussion} for additional discussion.

@@

 In the second approach, we leverage hospitalization information. We first estimate the probability of fever given demographic and COVID-19 test status.  When the COVID-19 test status is positive, the probability of fever is allowed to depend on whether the individual was hospitalized.  Figure~\ref{fig:symptomlik1_model2} and~\ref{fig:symptomlik2_model2} in Appendix~\ref{app:in_add_details} presents the estimated likelihood of reporting a fever with a COVID-19 positive individual given a negative and positive COVID-19 test in the past 24 hours respectively.  We see that hospitalization significantly increases the likelihood of fever as expected.  We use these models and the hospitalization rates by demographic to perform mean impuation of the number of positive and negative tests given demographic strata, fever status, and COVID-19 contact status. Inverse probability weights are then computed based on the resulting dataset (termed IPW2).

 To make the approach concrete, Figure~\ref{fig:testinglik_mainpaper} presents the testing propensity for non-hispanic, white males across age strata and fever status using the second approach.  We note here that the ratio of testing propensity within strata across fever status is time-varying.  The ratio starting 50 times more likely in early April and dropping to ten times as likely by end of 2020.

@@

 Here we consider the multinomial SEIR model using observed death data as presented in Figure~\ref{fig:in-deaths}. The model, presented in Section~\ref{section:modelbased}, requires infection fatality rates to be specified per-strata.  In this paper, similar to~\cite{Johndrow2020}, we specify fixed per-strata infection fatality rates. Based on published age-specific IFRs~\cite{Levin2020}, IFR closely follows a log-linear relationship with age. Combining across published age-specific IFRs, Table~\ref{tab:ifrperage} presents estimates of the average IFR within each strata using the age-distribution.
 \begin{table}[!th]
 \begin{tabular}{c | c c c c c c c c}
 Age Strata & 0-19 & 20-29 & 30-39 & 40-49 & 50-59 & 60-69 & 70-79 & 80+ \\ \hline
 IFR & 0.003\% & 0.014\% & 0.046\% & 0.144\% & 0.456\% & 1.443\% & 4.567\% & 15.468\%
 \end{tabular}
 \caption{IFR by age-strata}
 \label{tab:ifrperage}
 \end{table}
 The discrete-time distribution~$\{ \theta_{j} \}$ is a discretized version of a truncated normal distribution with mean $25$, standard deviation~$5$, minimum value $0$, and maximum value $44$.  This was chosen to closely mimic the distribution from~\cite{Johndrow2020}.

@@

 The SEIR model in equation~\eqref{eq:seir} is extended to allow for time-varying transmission rates~$\beta_t$ to account for public policy changes discussed in Section~\ref{section:publicpolicyindiana} over the time period considered.  Here, we insert three change points $(t_1,t_2,t_3)$ on March 23rd, June 15th, and October 1st of 2020 respectively.  Given sensitivity to these choices, instead of considering a multiplicative structure, we allow for delayed implementation and slow change in the transmission rate by using weights:
 $$
 \beta_t = \sum_{k=1}^{3} \left(\beta_{k-1} (1-w^{(k)}_t ) + \beta_{k} w_t\right),\quad w^{(k)}_t = (1 + \exp \left( -\xi (t - t_k - \nu) \right))^{-1}
 $$
 where $\nu \geq 0$ is the potential delay and $\xi$ is the rate of change.  Each weight~$w_t^{(k)} = 0$ for $t < t_k$ and $w_t^{(k)} \to 1$ as $t \to \infty$. See Section~\ref{app:prior_modelbased} in the supplementary materials for details on prior specification.

>>>>

\begin{abstract}
  Coronavirus case-count data has influenced government policies and drives most epidemiological forecasts. Limited testing is cited as the key driver behind minimal information on the COVID-19 pandemic. While expanded testing is laudable, measurement error and selection bias are the two greatest problems limiting our understanding of the COVID-19 pandemic; neither can be fully addressed by increased testing capacity. In this paper, we demonstrate their impact on estimation of point prevalence and the effective reproduction number. We show that estimates based on the millions of molecular tests in the US has the same mean square error as a small simple random sample.  To address this, a procedure is presented that combines case-count data and random samples over time to estimate selection propensities based on key covariate information. We then combine these selection propensities with epidemiological forecast models to construct a \emph{doubly robust} estimation method that accounts for both measurement-error and selection bias.  This method is then applied to estimate Indiana's prevalence using case-count, hospitalization, and death data with cumulative demographic information, a statewide random molecular sample collected from April 25--29th, and Facebook's COVID-19 symptom survey.  We end with a series of recommendations based on the proposed methodology.
\end{abstract}

@@

Due to monetary and time constraints, however, random testing is performed infrequently.  As of June 2021, Indiana and Ohio are the only states to conduct statewide random sample testing\footnote{These are is the only random samples to collect both seroprevalence and diagnostic testing results. The CDC and other states have conducted seroprevalence-only studies.}. Indiana's sample was collected from April 25--29, 2020~\citep{Yiannoutsos2021}.  Such infrequent random testing is likely to provide insufficient information to help researchers and policy makers better understand the disease trajectory which can change rapidly over time.  Therefore, while random testing may be preferable in theory, in practice governments, researchers, and policy makers continue to use coronavirus case-counts to understand the impact of COVID-19 on the population and make data-informed decisions.

@@

 This article discusses the relationship between three statistical concepts: selection bias, measurement error, and population size.  While the impact of measurement error~\citep{Smeden2019} and selection bias~\citep{Keiding2016} on estimation are both well-studied topics, here we provide a new perspective by building on the work of \cite{Meng2018} who studied an error decomposition to understand the relationship between selection bias and population size.  Specifically, we quantify the interaction between measurement-error and selection bias on statistical error, showing how the sign and magnitude can change drastically.
 % the impact on the effective sample size can therefore be quite large.

@@

 One core component of coronavirus research is epidemiological compartmental modelling of case-count and death data. These models can be used to answer a variety of research questions including case-count forecasting, estimation of the effective reproduction number, and estimation of quarantine and other health policies on infectious disease dynamics.  The basic approach is a deterministic compartmental model called the susceptible-infectious-recovered (SIR) model.  A probabilistic extension was proposed by~\cite{Osthus2017} to model one-dimensional time series of infected proportions. \cite{Song2020} extended this approach to incorporate interventions and assess interventions on COVID-19 epidemic in China. \cite{Hao2020} builds an extended SEIR model (SAPHIRE) to account for various presymptomatic infectiousness, time-varying ascertainment rates, transmission rates and population movements.

@@

 Between April 25--29, 2020, Indiana conducted statewide random molecular testing of persons ages $\geq 12$ years to assess prevalence of active infection to SARS-CoV-2~\citep{Yiannoutsos2021}. A stratified random sampling design was conducted using Indiana’s 10 public health preparedness districts as sampling strata. 15,495 participants were contacted resulting in a final sample size of 3,658. Demographic data was collected (e.g., summary statistics on age, sex, and race) as well as data on whether they experienced any COVID-19 compatible symptoms during the past 2 weeks or had shared a household with someone who had a positive test result for SARS-CoV-2. During May 2--3, 2020, an additional non-random sample of 898 individuals was also collected.
 % In this paper, the nonrandom sample is used to assess whether the cumulative summary statistics are appropriate in constructing the inverse-probability weights as well as in constructing appropriate weights using symptom information.

@@

 Here we assume access to a probability sample measuring the same set of covariates.  Let $\tilde I_j$ be an indicator that the individual was included in the probability sample and $\tilde W_j$ be the probability of inclusion for the $j$th individual (i.e., equal to $n/N$ for a simple random sample of size $n$). In recent work,~\cite{Chen2019} use a probability sample to construct a design-unbiased estimator of the second term
 \begin{equation}
 \label{eq:auxinfoprob}
 \sum_{j=1}^N I_j \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right)  + \sum_{i=1}^N \tilde I_j \tilde W_j \log ( 1 - \pi (X_j; \theta)).
 \end{equation}
 Taking an expectation with respect to the sampling design yields~\eqref{eq:propensity}. Solving~\eqref{eq:auxinfoprob} can be done by iteratively re-weighted least squares (IRLS); see Section~\ref{app:irls} in the supplementary materials for algorithmic details.

@@

 If sufficiently large random samples are collected at each time $t =1,\ldots,T$, then the pseudo-likelihood can be re-written as in~\eqref{eq:auxinfoprob} and propensities estimated separately per time point.  Unfortunately for COVID-19, large probabilistic samples are not available at every time within a given region.  Here we consider a non-parametric kernel-based approach where the selection propensity at time~$t$, denoted $\hat \theta_t$, maximizes the smoothed pseudo-likelihood
 $$
 \sum_{t^\prime=1}^T K_h(|t^\prime - t|) \left[ \sum_{j=1}^N I_{j,t^\prime} \log \left( \frac{\pi_t (X_{j,t^\prime}; \theta)}{1-\pi_t(X_{j,t^\prime}; \theta)} \right) + \sum_{j=1}^N \tilde W_{j,t^\prime} \tilde I_{j,t^\prime} \log \left( 1 - \pi_t (X_{j,t^\prime}; \theta) \right) \right]
 $$
 where $K_h$ is a kernel function with tunable parameter $h$. Given $\pi (x;\hat \theta_t)$, the prevalence estimator $\bar y_{n,t}^\star$ is given by~\ref{eq:ipwest} using case-count data observed on day~$t$.

@@

 \begin{lemma}[Variance of IPW estimator] \normalfont
 \label{lemma:ipw}
 % \textcolor{red}{Must check.}
 The estimates~$\hat \mu_t := \bar y_{n,t}^\star$, $\hat \pi_{j,t}$, $\hat FP$, and $\hat FN$ are solutions to the following set of estimating equations
 $$
 \Phi_n (\eta_t) =
 \left (
 \begin{array}{c}
 \frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - FP - (1-FP-FN) \cdot \mu_t}{\pi_{j,t}} \\
 \frac{1}{N} \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
 \frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
 \frac{1}{n_{FN}} \sum_{j \in S_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
 \end{array}
 \right ) = {\bf 0}
 $$
 where $\eta_t = (\mu_t, \pi_t, FP, FN)$. Under certain regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (\bar n^{-1/2})$  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (\bar n^{-1})$ where $\bar n = n \sum_{l=1}^L K_{t,t_l}$ and $V_t^{(IPW)}$ is the first diagonal element of $E [\Phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\Phi_n(\eta_0)]^{-1}$ with $E[\cdot]$ and $\text{Var} (\cdot)$ are under the joint randomization of propensity score and sampling designs.
 \end{lemma}

@@

 \subsection{Doubly robust estimation.}
 Rather than relying solely on IPW or epidemiological forecasts, here we combine forecasting and inverse-probability weighting by extending recent work by~\cite{Chen2019} to account time-varying propensities and measurement-error. In this context, the doubly-robust estimator is given by
 $$
 \bar y_{n}^{(DR)} = \frac{1}{N} \sum_{j=1}^N \hat \mu (x_j) + \frac{1}{\sum_{j=1}^N I_j w (x_j)} \sum_{j=1}^N I_j w(x_j) \left( \frac{Y_j - FP}{1 - FP - FN} - \mu(x_j) \right).
 $$
 where the estimates $\hat \mu(x_j)$ are not corrected for measurement-error as they are built to estimate true case count from death-only data. This estimator is called ``doubly-robust'' because it is consistent if either the model-based forecasts or the time-varying propensities are correctly specified. A statistical error decomposition can be derived
 \begin{align}
 \label{eq:statdecomp3}
 \begin{split}
 \rho_{\tilde I (X), Y-\mu(X)} &\times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y-\mu(X)}  \\
 &\times \underbrace{\left[ 1 - \frac{\rho_{\tilde I(X),Y} \sigma_{Y}}{\rho_{\tilde I(X),Y-\mu(X)} \sigma_{Y-\mu(X)}} \times \tilde \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{\tilde f_0 (1-\bar Y) + \tilde f_1 \bar Y} \right]}_{\tilde D_M}.
 \end{split}
 \end{align}
 See Appendix~\ref{app:DRderivation} for the derivation.
 % Comparing~\eqref{eq:statdecomp3} to~\eqref{eq:statdecomp2} shows that the addition of the model-based forecast impacts the estimation error in two ways.
 If the model is adequate, then one may expect a reduction in the problem difficulty and (potentially) in the data quality components.
 % Note that the model-based forecast does not impact the data quantity component.
% Second, if the model-based forecast is correct then $E_{\bf R} [ \rho_{\tilde I (X), Y-\mu(X)} ] = 0$  and therefore $\rho_{\tilde I(X), Y-\mu(X)} = O(N^{-1})$; however, if the forecasts are not correctly specified then the data quality index is unlikely to inversely scale with population size.
 Interestingly, the impact of measurement-error on data quality now depends on a relative comparison of the data quality and problem difficulty of the weighted estimator and the doubly-robust estimator.
 % For decent models, one would expect this ratio to be greater than one.  Therefore, $\tilde \Delta < 0$ implies an expected increase in impact of measurement-error and $\tilde \Delta > 0$ implies the magnitude could shrink toward zero or change sign and have an even large magnitude in the opposite direction.  Again, if the data quality is not at the level of $N^{-1}$, then confidence intervals constructed from a DR estimator are likely to put too much confidence in the sheer size of the data.

@@

 \begin{lemma}[Variance of doubly-robust estimator] \normalfont
 \label{lemma:dr}
 The estimates~$\mu_t := \bar y_{n,t}^\star$ and $\hat \pi_{j,t}$
 are solutions to the following set of estimating equations
 $$
 \Phi_n (\eta_t) =
 \left (
 \begin{array}{c}
 \frac{1}{N} \left[ \sum_{j=1}^N \frac{I_{j,t}}{\pi_{j,t}} \left( \left( \frac{Y_{j,t} - FP}{1-FP-FN} - \hat \mu_{j,t}  \right)  - \mu_t \right) + \frac{1}{\sum_{j=1}^N \tilde I_j \tilde W_j} \sum_{j=1}^N  \tilde I_j \tilde W_j \hat \mu_t (X_{j,t}) \right] \\
 \frac{1}{N} \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
 \frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
 \frac{1}{n_{FN}} \sum_{j \in S_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
 \end{array}
 \right ) = {\bf 0}
 $$
 where $\eta_t = (\mu_t, \pi_t, FP, FN)$. Under regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (\bar n^{-1/2})$ and $\text{var} (\hat y_{n,t}) = V_{t}^{(DR)} + o (\bar n^{-1})$ where $\bar n = n \sum_{l=1}^L K_{t,t_l}$ and $V_t^{(DR)}$ is the first diagonal element of $E [\Phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\Phi_n(\eta_0)]^{-1}$ with $E[\cdot]$ and $\text{Var} (\cdot)$ are under the joint randomization of propensity score and sampling designs.
 \end{lemma}

@@

 \item {\bf Random/Nonrandom statewide sample}: Table~\ref{tab:indiana} summarizes data from a random sample from April 25--29th as well as a nonrandom sample obtained between May 2--3 in racial/ethnic minority communities~\cite{Yiannoutsos2021}. The nonrandom sample is not a random subsample of the overall case-counts; however, this data provides important supplementary covariate information.
 % Namely, the samples collected included covariate information on (1) prior symptoms, (2) COVID-19 positive individual in household, and (3) prior COVID-19 positive test.
 To account for the nonrandom sample being from high risk areas, we adjust estimates to match the statewide positivity rate of 11.7\% on August 30th. These adjusted rates are presented in parentheses in Table~\ref{tab:indiana} along with the relevant covariate information.

@@

 The IPW analysis is next extended to the time-varying setting.  To do so, we make use of the Facebook symptom survey, hospitalization, and testing data.  Recall the COVID-19 testing and positive case counts are known by age, gender, and racial demographics. Symptom status (e.g., fever, coughing, shortness of breath) and related covariate information (COVID-19 positive contact), however, are currently unavailable\footnote{Whether this is due to data privacy considerations or lack of data collection efforts is unclear.}.  As COVID-19 symptom status is likely to alter the testing propensity and likelihood of active infection, for illustrative purposes Facebook symptom survey and hospitalization data are used in two ways to impute symptom status per strata. We outline the imputation methods below; see Section~\ref{section:discussion} for additional discussion.

@@

 In the second approach, we leverage hospitalization information. We first estimate the probability of fever given demographic and COVID-19 test status.  When the COVID-19 test status is positive, the probability of fever is allowed to depend on whether the individual was hospitalized.  Figure~\ref{fig:symptomlik1_model2} and~\ref{fig:symptomlik2_model2} in Appendix~\ref{app:in_add_details} presents the estimated likelihood of reporting a fever with a COVID-19 positive individual given a negative and positive COVID-19 test in the past 24 hours respectively.  We see that hospitalization significantly increases the likelihood of fever as expected.  We use these models and the hospitalization rates by demographic to perform mean imputation of the number of positive and negative tests given demographic strata, fever status, and COVID-19 contact status. Inverse probability weights are then computed based on the resulting dataset (termed IPW2).

 Figure~\ref{fig:testinglik_mainpaper} presents the testing propensity for non-Hispanic, white males across age strata and fever status using the second approach.  Note that the ratio of testing propensity within strata across fever status is time-varying, starting 50 times more likely in early April and dropping to ten times more likely by end of 2020.

@@

 Here we consider the multinomial SEIR model using observed death data as presented in Figure~\ref{fig:in-deaths}. The model, presented in Section~\ref{section:modelbased}, requires infection fatality rates to be specified per-strata.  In this paper, similar to~\cite{Johndrow2020}, we specify fixed per-strata infection fatality rates.  Based on published age-specific IFRs~\cite{Levin2020}, IFR closely follows a log-linear relationship with age. \cite{Ironse2103272118} uses Indiana's random survey and death data to estimate a marginal IFR of $0.84\%$.  Combining across published age-specific IFRs and anchoring our analysis to the estimated marginal IFR of $0.84\%$ for Indiana, Table~\ref{tab:ifrperage} presents the age-specific IFRs used in the analysis. Note we collapse ages less than $40$ given the limited number of COVID-19 related deaths in this age range.
 \begin{table}[!th]
 \begin{tabular}{c | c c c c c c c c}
 Age Strata & 0-40 & 50-59 & 60-69 & 70-79 & 80+ \\ \hline
 IFR & $0.014\%$ & $0.120\%$ & $1.206\%$ & $3.815\%$ & $12.920\%$
 \end{tabular}
 \caption{IFR by age-strata}
 \label{tab:ifrperage}
 \vspace{-1cm}
 \end{table}
 The discrete-time distribution~$\{ \theta_{j} \}$ is a discretized version of a truncated normal distribution with mean $25$, standard deviation~$5$, minimum value $0$, and maximum value $44$.  This was chosen to closely mimic the distribution from~\cite{Johndrow2020}.

@@

 The SEIR model in equation~\eqref{eq:seir} is extended to allow for time-varying transmission rates~$\beta_t$ to account for public policy changes discussed in Section~\ref{section:publicpolicyindiana} over the time period considered.  Here, we insert three change points $(t_1,t_2,t_3)$ on March 23rd, June 15th, and October 1st of 2020 respectively.  Given sensitivity to these choices, instead of considering a multiplicative structure, we allow for delayed implementation and slow change in the transmission rate by using weights:
 $$
 \beta_t = \sum_{k=1}^{3} \left(\beta_{k-1} (1-w^{(k)}_t ) + \beta_{k} w_t\right),\quad w^{(k)}_t = (1 + \exp \left( -\xi (t - t_k - \nu) \right))^{-1} {\bf 1}[t \geq t_k]
 $$
 where $\nu \geq 0$ is the potential delay, $\xi$ is the rate of change, and ${\bf 1}[\cdot]$ is an indicator function.  Each weight~$w_t^{(k)} = 0$ for $t < t_k$ and $w_t^{(k)} \to 1$ as $t \to \infty$. See Section~\ref{app:prior_modelbased} in the supplementary materials for details on prior specification.
